Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-12-08 paper/abstract.tex
letterizing

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
proxy, to reveal their rotational evolution and demonstrate that kinematics
could be a useful tool for calibrating gyrochronology.
At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
to the Praesepe cluster, accurately predicts the relative ages of stars.
At these young ages, lower-mass stars spin more slowly than higher-mass stars
because their stronger magnetic fields lead to more efficient angular momentum
loss.
However, at old ages we find that late G and early K dwarfs rotate {\it
faster} than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results, based on field star rotation periods, align with recent
findings from the rotation periods of stars in middle-aged open clusters and
theoretical models that vary the rate of surface-to-core angular momentum
transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
proxy, to reveal their rotational evolution and demonstrate that kinematics
could be a useful tool for calibrating gyrochronology.
% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
% to the Praesepe cluster, accurately predicts the relative ages of stars.
We find that, at young ages, lower-mass stars spin more slowly than
higher-mass stars because their stronger magnetic fields lead to more
efficient angular momentum loss.
This reflects the behavior of young open clusters such as the Pleaides,
Praesepe, and the Hyades.
However, at older ages we find that late G and early K dwarfs rotate at the
same rate as late K dwarfs of the same age, and eventually rotate faster.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results align with recent findings from the rotation periods of stars in
middle-aged open clusters and theoretical models that vary the rate of
surface-to-core angular momentum transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

### 2019-12-08 paper/discussion.tex
letterizing

The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass.
This is probably because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfven radii and therefore experience greater
angular momentum loss rate.
According to the \citet{spada2019} models, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for GK dwarfs.
The explanation for this, according to the \citet{spada2019} models, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two interior layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly.

>>>>

The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass, likely because
lower-mass stars with deeper convection zones have stronger magnetic fields,
larger Alfv\'en radii and therefore experience greater angular momentum loss
rate.
According to the \citet{spada2019} models, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for GK dwarfs.
The explanation for this, according to the \citet{spada2019} models, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two interior layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly.

### 2019-12-08 paper/intro.tex
letterizing

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and spin more slowly over time through magnetic braking
\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are typically born with random rotation periods, ranging from 1
to 10 days, observations of young open clusters reveal that their rotation
periods converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009}.
After this time, it is thought that the rotation period of a star is
determined, to first order, by its color and age alone.
This is the principle behind gyrochronology -- the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
% It is well established that magnetized stellar winds cause the rotation
% periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
% weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
% calibrated, the relationship between period, mass and age can be used to
% age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
% meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} reveal that rotational evolution is more complicated
than previously thought.
For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
slowly than the G dwarfs -- in theory because lower-mass stars have deeper
convections zones which generate stronger magnetic fields and more efficient
magnetic braking.
However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
same} rate as early-K dwarfs \citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period-color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.

Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (ratio of rotation period to
convective turnover time) of around 2 \citep{vansaders2016, vansaders2018}.

The rotational evolution of stars is clearly a complicated process, and to
fully calibrate the gyrochronology relations we need reliable ages for large
numbers of field stars which span a range of ages.
% however relying on open clusters (which are typically young) and asteroseismic
% stars (which are typically G-type stars or more massive) to calibrate
% gyrochronology relations limits the mass and age coverage of the calibration
% sample.
In this paper, we use the velocity dispersions of field stars to gain insight
into their rotational evolution.
Velocity dispersion provides relative ages for groups of stars, allowing us to
observe the rotational evolution of stars over a range of masses and ages.

@@

% Vertical action is thought to be a better age indicator than vertical
% velocity \citep{ting2019}, although still only a weak age indicator for
% individual stars \citep{beane2018}, however both vertical action ($J_z$) and
% vertical
Vertical velocity, \vz, can only be calculated with full 6-dimensional
position and velocity information, and unfortunately most stars with measured
rotation periods do not have radial velocity (RV) measurements because they
are relatively faint \kepler\ targets ($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb,
to approximate \vz.
The \kepler\ field is positioned at low galactic latitude
(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect -- see section
\ref{sec:results}) approximation to \vz.
Because we use \vb\ rather than \vz\, we cannot calculate absolute kinematic
ages using an age-velocity dispersion relation (AVR).
However, regardless of direction, velocity dispersion is expected to
monotonically increase over time, and can therefore be used to {\it rank}
groups of stars by age.

>>>>

Stars with significant convective envelopes ($\lesssim$ 1.3 M$_\odot$) have
strong magnetic fields and spin more slowly over time through magnetic braking
\citep[\eg][]{schatzman1962, weber1967, skumanich1972, kawaler1988,
pinsonneault1989}.
Although stars are typically born with random rotation periods, ranging from 1
to 10 days, observations of young open clusters reveal that their rotation
periods converge onto a unique sequence by $\sim$500-700 million years
\citep[\eg][]{irwin2009, gallet2013}.
After this time, the rotation period of a star is thought to be determined, to
first order, by its color and age alone.
This is the principle behind gyrochronology -- the method of inferring a
star’s age from its rotation period \citep[\eg][]{barnes2003, barnes2007,
barnes2010, meibom2011, meibom2015}.
% It is well established that magnetized stellar winds cause the rotation
% periods of FGKM dwarfs to increase over time \citep[\eg][]{schatzman1962,
% weber1967, skumanich1972, kawaler1988, pinsonneault1989}, and, once fully
% calibrated, the relationship between period, mass and age can be used to
% age-date stars via gyrochronology \citep{barnes2003, barnes2007, barnes2010,
% meibom2011, meibom2015}.
However, new photometric rotation periods made available by the \kepler\
\citep{borucki2010} and \ktwo\ \citep{howell2014} missions
\citep[\eg][]{mcquillan2014, garcia2014, douglas2017, rebull2017, meibom2011,
meibom2015, curtis2019} reveal that rotational evolution is more complicated
than previously thought.
For example, the M dwarfs in the $\sim$ 650 Myr Praesepe cluster spin more
slowly than the G dwarfs.
In theory this is because lower-mass stars have deeper convections zones which
generate stronger magnetic fields and more efficient magnetic braking.
However, in the 1.1 Gyr NGC 6811 cluster, late-K dwarfs rotate at the {\it
same} rate as early-K dwarfs \citep{curtis2019}.
In other words, convection zone depth cannot be the only variable that affects
stellar spin-down rate.
New semi-empirical models that vary the rate of angular momentum
redistribution in the interiors of stars are able to reproduce this flattened
period-color relation \citep{spada2019}.
These models suggest that mass and age-dependent angular momentum transport
between the cores and envelopes of stars has a significant impact on their
surface rotation rates.
Another example of unexpected rotational evolution is seen in old field stars
which appear to rotate more rapidly than classical gyrochronology models
predict \citep{angus2015, vansaders2016, vansaders2018, metcalfe2019}.
A mass-dependent modification to the classical \prot\ $\propto
t^{\frac{1}{2}}$ spin-down law \citep{skumanich1972} is required to reproduce
these observations.
To fit magnetic braking models to these data, a cessation of magnetic braking
is required after stars reach a Rossby number (Ro; the ratio of rotation
period to convective turnover time) of around 2 \citep{vansaders2016,
vansaders2018}.

The rotational evolution of stars is clearly a complicated process and, to
fully calibrate the gyrochronology relations we need a large sample of
reliable ages for stars spanning a range of ages and masses.
% however relying on open clusters (which are typically young) and asteroseismic
% stars (which are typically G-type stars or more massive) to calibrate
% gyrochronology relations limits the mass and age coverage of the calibration
% sample.
In this paper, we use the velocity dispersions of field stars to qualitatively
explore the rotational evolution of GKM dwarfs, and show that kinematics could
provide a gyrochronology calibration sample.

@@

% Vertical action is thought to be a better age indicator than vertical
% velocity \citep{ting2019}, although still only a weak age indicator for
% individual stars \citep{beane2018}, however both vertical action ($J_z$) and
% vertical
Vertical velocity, \vz, can only be calculated with full 6-dimensional
position and velocity information, and unfortunately most stars with measured
rotation periods do not have radial velocity (RV) measurements because they
are relatively faint \kepler\ targets ($\sim$12th-17th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb,
to approximate \vz.
The \kepler\ field is positioned at low galactic latitude
(b=$\sim$5-20\degrees), so \vb\ is a close (although imperfect -- see section
\ref{sec:results}) approximation to \vz.
Because we use \vb\ rather than \vz\, we cannot calculate absolute kinematic
ages using an age-velocity dispersion relation (AVR).
However, regardless of direction, velocity dispersion is expected to
monotonically increase over time, and can therefore be used to {\it rank}
groups of stars by age.

### 2019-12-08 paper/method.tex
letterizing

At a galactic latitude of zero, $v_b=v_z$, however for increasing values of
$b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex.
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with a standard deviation (calculated as 1.5 $\times$ the median absolute
deviation) that increased with increasing $b$.
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms${^-1}$ at 10$^\circ$, 6kms${^-1}$ at 15$^\circ$, and 9kms${^-1}$ at
20$^\circ$.
Since we are concerned with velocity {\it dispersions}, rather than velocities
themselves, we also compared the \vb\ and \vz\ velocity dispersions for stars
downloaded from the GUMS simulation.
For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
latitudes the two quantities became significantly different.
For this reason we proceeded by only including stars with galactic latitudes
less than 15\degrees\ in our analysis.

@@

>>>>

At a galactic latitude of zero, $v_b=v_z$, however for increasing values of
$b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
Galaxy.
We subtracted this offset and examined the residuals of the \vz\-\vb\
relationship to investigate the variance as a function of Galactic latitude.
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with a standard deviation (calculated as 1.5 $\times$ the median absolute
deviation) that increased with increasing $b$.
Figure \ref{fig:vb_vz} shows how \vz\ transforms to \vb\ over a range of
Galactic latitudes.
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms${^-1}$ at 10$^\circ$, 6kms${^-1}$ at 15$^\circ$, and 9kms${^-1}$ at
20$^\circ$.
Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
over different Galactic latitudes.
Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
fits are shown as dashed blue lines.
The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
the variance increases with increasing Galactic latitude.
\begin{figure}
  \caption{
This figure demonstrates the relationship between \vb\ and \vz\, after
    subtracting a straight-line-fit, for stars in
    the \kepler\ field, based on the GUMS simulation.
The panels show a kernel density estimator (KDE) (black solid line) for
    the \vz-\vb\ residuals of stars in the GUMS simulation at four different
    Galactic latitudes.
Blue dashed lines show Gaussian fits to these KDEs.
The distributions are close to Gaussians, but with slightly heavy tails.
The standard deviations of the Gaussian fits increase with Galactic latitude.
This figure illustrates the fact using \vb\ instead of \vz\ artificially
    increases the velocity dispersion, especially at high latitudes.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vz}
\label{fig:vb_vz}
\end{figure}

Since we are concerned with velocity {\it dispersions}, rather than velocities
themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
function of temperature for
stars downloaded from the GUMS simulation.
For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
latitudes the two quantities became significantly different.
For this reason we proceeded by only including stars with galactic latitudes
less than 15\degrees\ in our analysis.
Although it seems that the transformation between \vz\ and \vb\ does not {\it
strongly} affect our results, we cannot rule out the possibility that it
introduces systematic biases into the velocity dispersions we present here.
In \gaia\ DR3, RVs will be available for most stars in this sample, providing
an opportunity to validate (or correct) the results presented here.

@@

In this paper, we do not attempt to convert velocity dispersion (\sigmavb)
into an age via an age-velocity dispersion relation (AVR)
\citep[\eg][]{holmberg2009}.
Although it seems \sigmavb\ can be used to roughly rank stars by age, a more
careful analysis will be needed to calculate absolute ages.
For example, the velocity distributions could be modeled as a mixture of
Gaussians in order to account for the additional velocity dispersion caused by
the \vz-\vb\ transformation.
The RVs of most of these stars will become available in \Gaia\ DR3, allowing
calculations of \vz, which can be used to calculate more reliable ages via an
AVR.
